---
title: "Project 1 - Maab Aldulimy"
format: html
editor: visual
---

# Project Goal

First Project in ST558 the goal is to write functions that will manipulate and process data sets that come from a census API. The generic functions should automatically summarize and plot certain returned data. Lastly, a document via quarto will be written to describe the thought process and give examples of using the functions.

## Data

The data for this project will be acquired from the Public Use Microdata Sample (PUMS) Census API. This API gives pseudo person level data. An API was requested and acquired from the website

```{r}
# loading in needed packages 

library(dplyr)
library(httr)
library(jsonlite)
library(ggplot2)

# masking API
api_key <- Sys.getenv("CENSUS_API_KEY")

if (api_key == "") {
  stop("CENSUS_API_KEY has not been set.")
}
```

##Part 1

For this section, we connect to the census API

We use GET() from the httr package to return information about the variables in the census. our aim at this point is to make sure the URL works and that we can access the information we want.In the code below we try to access the items under the variable sex.

```{r}
url <- paste0("https://api.census.gov/data/2024/acs/acs1/pums/variables.json")

variables_f <- GET(url)
vars_2024 <- fromJSON(content(variables_f,"text"))

#testing to see if i can access male/female 
vars_2024$variables$SEX$values$item


```

Below we make a character vector of the 15 variable names ( AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP, PWGTP, FER, HHL, SCH, SCHL, SEX, ST (for 2021, 2022)/STATE (for 2023, 2024), REGION, and DIVISION) and subset vars_2024\$variables with single brackets.

```{r}
 vars_2024_sub <- vars_2024$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST","STATE","REGION","DIVISION")]

names(vars_2024_sub)
length(vars_2024_sub)
```

The above code chunk ran, but one of the variable names came back as NA and the list had 16 elements instead of 15. This is expected: the project instructions say the state variable is called ST in 2021 and 2022 and STATE in 2023 and 2024. Since I'm querying the 2024 data, there is no ST variable, so R returns an empty NA element for it.

One easy fix would be to define a different set of wanted variables for each group of years (leaving out ST for 2023/2024 and STATE for 2021/2022). However, since my goal is to build a function that pulls this information automatically, a more general solution is better. In the code chunk below, I include both ST and STATE in wanted_vars, then use %in% to keep only the names that actually appear in the 2024 dictionary.

```{r}
wanted_vars<-c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST","STATE","REGION","DIVISION")

#In the code below you can see that position 13 (ST) is FALSE.
wanted_vars %in% names(vars_2024$variables)
vars_to_keep <- wanted_vars[wanted_vars %in% names(vars_2024$variables)]


# redefining vars_2024_sub to not have the NA value 
vars_2024_sub <- vars_2024$variables[vars_to_keep]
names(vars_2024_sub)
length(vars_2024_sub)
```

Now that I have my 15 variables without any NAs I want to extract just the values of each of these 15 lists. In order to do this I will use the function lapply().

```{r}
vars_2024_values <- lapply(vars_2024_sub, function(x) x$values)

#testing if the above code worked correctly 
vars_2024_values$SEX
```

### Making a Function to get the dictionary

The project wants one list per year (2021 to 2024), and the only thing that changes between years is the year inside the URL.This means I need to build a function that can build the URL differently for each year. The paste0() function can be used for this so that only the year can be changed in the URL.The argument for my function is the year since its the only changing variable. I then can use all my previous code chunks just renamed to a more general form to extract the information from any given year that is run to the function.In essence the function will take a year argument then will builds the URL with paste0() runs my steps, and return a vars_all_values list. The wanted_vars can stay outside the function since it's the same for every year.

I run a test to see if the function will work with get_var_values(2021). I first check the names to see if it include ST and then I check that \$SEX still gives Male/Female.

```{r}

wanted_vars <- c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","ST","STATE","REGION","DIVISION")

get_var_values <- function(year) {
  url <-  paste0("https://api.census.gov/data/",
                 year,
                 "/acs/acs1/pums/variables.json")
  variables_m <- GET(url)
  vars_all <- fromJSON(content(variables_m, "text"))
  vars_to_keep <- wanted_vars[wanted_vars %in% names(vars_all$variables)]
  vars_all_sub <- vars_all$variables[vars_to_keep]
  vars_all_values <- lapply(vars_all_sub, function(x)
    x$values)
  return(vars_all_values)
}

#test run with 2021 data 
test_2021 <- get_var_values(2021)
names(test_2021)
test_2021$SEX
```

Now I need to Calling the function for all four years. I could call it four times, but since we are trying to make a pipeline I wanted to try and use use lapply() over the vector of years c(2021, 2022, 2023, 2024). This is nice because it allows us to easily change the data as needed.Meaning if I wanted data from 2000-2005 I can simply change the year vector and the function will retrieve the data from those years for me.

```{r}
year_vector <- c("2021" = 2021, "2022" = 2022,"2023" = 2023,"2024" = 2024)
comb_list <- lapply(year_vector,get_var_values)
names(comb_list)

#Test  
comb_list$`2021`$SEX
names(comb_list$`2021`)
names(comb_list$`2024`)

#saving data as .rds
saveRDS(comb_list, "data/comb_list.rds")
```

```{r}
url1 <- paste0("https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&for=state:*&SCHL=24&key=",api_key)

variables_y <- GET(url1)
parsed_data <- fromJSON(content(variables_y,"text"))

#testing 

class(parsed_data)
dim(parsed_data)
parsed_data[1:3, ]


variables_y$status_code
substr(content(variables_y, "text"), 1, 600)

```

Pull out row 1 to use as column names using parsed_data\[1, \]. Drop row 1 from the data using negative indexing parsed_data\[-1, \] which should give "everything except row 1."Then Convert to a tibble with as_tibble()

```{r}
col_names <- parsed_data[1, ]
colnames(parsed_data) <- col_names
data_tibble <- as_tibble(parsed_data[-1,])
head(data_tibble)
```

creating the helper tibble function

```{r}
helper_tibble <- function(resp) {
  parsed_data <- fromJSON(content(resp,"text"))
  col_names <- parsed_data[1, ]
  colnames(parsed_data) <- col_names
  data_tibble <- as_tibble(parsed_data[-1,])
  return(data_tibble)
}

#test 
test_tibble <- helper_tibble(variables_y)
head(test_tibble)
```

### Building a query function to handle year and filter requirements

The variables of the function will be year (default 2024), the numeric variables (default c("AGEP", "PWGTP")), the categorical variables (default "SEX"), the geography level (default "State"), the specific region/division/state to subset on (default is NC which is state 37).In order for this to work we will have to test each layer before moving on to the next, so if something breaks, we know where the issue is and can fix it.

```{r}
get_census_data <- function(year = 2024,
                             num_vars = c("AGEP", "PWGTP"),
                             cat_vars = "SEX",
                             geo_level = "State",
                             geo_subset = NULL) { 
  
  vars_string <- paste(c(num_vars, cat_vars), collapse = ",")
  url <- paste0("https://api.census.gov/data/",
              year,
              "/acs/acs1/pums?get=",
              vars_string,
              "&for=state:37&key=",
              api_key) 
  
  data_rsp <- GET(url)
  data_all_tibble <- helper_tibble(data_rsp)
  return(data_all_tibble)
}

```

#### First Test: Checking to see the function works

```{r}
# test if this works 
test_single <- get_census_data()
head(test_single)

```

Adding a year argument

```{r}
 get_census_data <- function(year = 2024,
                             num_vars = c("AGEP", "PWGTP"),
                             cat_vars = "SEX",
                             geo_level = "State",
                             geo_subset = NULL) { 
  
  vars_string <- paste(c(num_vars, cat_vars), collapse = ",")
  
# check that a single valid year was given
  if (length(year) != 1) {
    stop("you can only give one year at a time")
  }
  if (!(year %in% 2021:2024)) {
    stop("year must be 2021, 2022, 2023, or 2024")
  }
  url <- paste0("https://api.census.gov/data/",
              year,
              "/acs/acs1/pums?get=",
              vars_string,
              "&for=state:37&key=",
              api_key) 
  
  data_rsp <- GET(url)
  data_all_tibble <- helper_tibble(data_rsp)
  return(data_all_tibble)
}
```

#### Second Test: Checking to see that the year argument works

Below I test the year checks by asking for 2019 (outside the allowed range) and for two years at once. Both should return my error messages. Then I run the function with a valid year to make sure it still pulls data.

```{r}
#| error: true
# to prevent rendering issue we include the error:true line
get_census_data(year = 2019) # should give error because 2019 is not withing the specified year range 
```

```{r}
#| error: true
# to prevent rendering issue we include the error:true line
get_census_data(year = c(2022, 2023)) # should give error
```

```{r}
get_census_data(year = 2023) # should work 
```

Adding a numeric and categorical variables argument

```{r}
 get_census_data <- function(year = 2024,
                             num_vars = c("AGEP", "PWGTP"),
                             cat_vars = "SEX",
                             geo_level = "State",
                             geo_subset = NULL) { 
  
### check that a single valid year was given##############################
  if (length(year) != 1) {
    stop("you can only give one year at a time")
  }
  if (!(year %in% 2021:2024)) {
    stop("year must be 2021, 2022, 2023, or 2024")
  }
  
###check numeric variables################################################
   allowed_num_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  
   if (!all(num_vars %in% allowed_num_vars)) {
    stop(paste("numeric variables must be from:",
               paste(allowed_num_vars, collapse = ", ")))
  }
  if (!("PWGTP" %in% num_vars)) {
    stop("PWGTP must always be included")
  }
  if (length(setdiff(num_vars, "PWGTP")) < 1) {
    stop("you must include at least one numeric variable besides PWGTP")
  }
  
###check categorical variables###########################################
  allowed_cat_vars <- c("FER", "HHL", "SCH", "SCHL", "SEX") 
  
  if (length(cat_vars) < 1) {
    stop("you must include at least one categorical variable")
  }
  if (!all(cat_vars %in% allowed_cat_vars)) {
    stop(paste("categorical variables must be from:",
               paste(allowed_cat_vars, collapse = ", ")))
  }
  
###Build the URL and query the API ######################################
  
  vars_string <- paste(c(num_vars, cat_vars), collapse = ",")  
  url <- paste0("https://api.census.gov/data/",
              year,
              "/acs/acs1/pums?get=",
              vars_string,
              "&for=state:37&key=",
              api_key) 
  
  data_rsp <- GET(url)
  data_all_tibble <- helper_tibble(data_rsp)
  return(data_all_tibble)
}

```

#### Third Test: Checking the numeric variables

The numeric variable argument has several requirements. Only the numeric variables specified in the project instructions should be accepted. PWGTP must always be included because it represents the weight of each observation, and at least one additional numeric variable must also be included.The first check catches typos or variables outside the allowed list before they are sent to the API. The PWGTP check matters because each row stands for a different number of people, so the summaries and plots later in this project depend on having the weights. The last check makes sure there is something to actually analyze, since PWGTP on its own is only a weight.

The categorical variable argument follows a similar pattern. At least one categorical variable must be given, because the plotting function later in this project uses it to group the data, and the summary function counts its levels. Every categorical variable must also come from the list in the project instructions (FER, HHL, SCH, SCHL, and SEX). 

```{r}
#| error: true
# to prevent rendering issue we include the error:true line
get_census_data(num_vars = c("AGEP", "PWGTP", "HEIGHT"))
```

```{r}
#| error: true
# to prevent rendering issue we include the error:true line
get_census_data(num_vars = "AGEP")

```

```{r}
#| error: true
# to prevent rendering issue we include the error:true line
get_census_data(num_vars = "PWGTP")
```

```{r}
#| error: true
# to prevent rendering issue we include the error:true line
get_census_data(cat_vars = c("SEX", "MAR"))
```

```{r}
#| error: true
# to prevent rendering issue we include the error:true line
get_census_data(cat_vars = character(0))
```
```{r}
get_census_data(num_vars = c("AGEP", "GASP", "PWGTP"), cat_vars = c("SEX", "HHL"))
```

Adding the geography layer to the function :

Next I add the geography arguments. The user can choose one geography level (Region, Division, or State) and which specific area to pull data for. If no area is given, the function defaults to North Carolina, or the division and region that include it, which keeps the default query small and fast. The function also checks that the area codes are valid using the variable values I saved earlier in the .rds file, and the user can enter "all" to get every area at that level.

```{r}
 get_census_data <- function(year = 2024,
                             num_vars = c("AGEP", "PWGTP"),
                             cat_vars = "SEX",
                             geo_level = "State",
                             geo_subset = NULL) { 
  
### check that a single valid year was given##############################
  if (length(year) != 1) {
    stop("you can only give one year at a time")
  }
  if (!(year %in% 2021:2024)) {
    stop("year must be 2021, 2022, 2023, or 2024")
  }
  
###check numeric variables################################################
   allowed_num_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  
   if (!all(num_vars %in% allowed_num_vars)) {
    stop(paste("numeric variables must be from:",
               paste(allowed_num_vars, collapse = ", ")))
  }
  if (!("PWGTP" %in% num_vars)) {
    stop("PWGTP must always be included")
  }
  if (length(setdiff(num_vars, "PWGTP")) < 1) {
    stop("you must include at least one numeric variable besides PWGTP")
  }
  
###check categorical variables###########################################
  allowed_cat_vars <- c("FER", "HHL", "SCH", "SCHL", "SEX") 
  
  if (length(cat_vars) < 1) {
    stop("you must include at least one categorical variable")
  }
  if (!all(cat_vars %in% allowed_cat_vars)) {
    stop(paste("categorical variables must be from:",
               paste(allowed_cat_vars, collapse = ", ")))
  }

 ### check geography level ###
  allowed_geo_levels <- c("Region", "Division", "State")

  if (length(geo_level) != 1) {
    stop("you can only choose one geography level")
  }
  if (!(geo_level %in% allowed_geo_levels)) {
    stop(paste("geo_level must be one of:",
               paste(allowed_geo_levels, collapse = ", ")))
  }

  ### set default subset if none given (NC and the areas it belongs to) ###
  if (is.null(geo_subset)) {
    if (geo_level == "State") {
      geo_subset <- "37"
    } else if (geo_level == "Division") {
      geo_subset <- "5"
    } else {
      geo_subset <- "3"
    }
  }

  ### check subset values against the saved variable values ###
  var_values <- readRDS("data/comb_list.rds")[[as.character(year)]]

  if (geo_level == "State") {
    geo_var <- ifelse("STATE" %in% names(var_values), "STATE", "ST")
  } else {
    geo_var <- toupper(geo_level)
  }
  allowed_codes <- names(var_values[[geo_var]]$item)

  if (!(length(geo_subset) == 1 && tolower(geo_subset) == "all")) {
    if (!all(geo_subset %in% allowed_codes)) {
      stop(paste("invalid", geo_level, "code. Allowed codes are:",
                 paste(allowed_codes, collapse = ", ")))
    }
  }

  ### turn the subset into the URL format ###
  if (length(geo_subset) == 1 && tolower(geo_subset) == "all") {
    geo_string <- "*"
  } else {
    geo_string <- paste(geo_subset, collapse = ",")
  }

  ### build the URL and query the API ###
  vars_string <- paste(c(num_vars, cat_vars), collapse = ",")
  url <- paste0("https://api.census.gov/data/",
                year,
                "/acs/acs1/pums?get=",
                vars_string,
                "&for=", tolower(geo_level), ":", geo_string,
                "&key=",
                api_key)
   
  data_rsp <- GET(url)
  data_all_tibble <- helper_tibble(data_rsp)
  return(data_all_tibble)
 }
```

####Fourth Test: Checking to see if the geography checks work

```{r}
#| error: true
get_census_data(geo_level = c("State", "Region"))   # error: one level only
```

```{r}
#| error: true
get_census_data(geo_level = "County")               # error: not allowed
```
```{r}
#| error: true
get_census_data(geo_subset = "99")                  # error: invalid code
```

```{r}

get_census_data()                                   # NC, should work
get_census_data(geo_level = "Region")               # South, should work
get_census_data(geo_level = "State", geo_subset = c("37", "51"))  # NC and VA

```

Next we convert the numeric variables into the right types, since every column currently comes back from the API as text. Most of them (AGEP, GASP, GRPIP, JWMNP, and PWGTP) only need as.numeric().The two time variables, JWAP (time of arrival at work) and JWDP (time of departure for work), are harder because the API returns a code instead of a time. For example, a JWAP code of 1 means "12:00 a.m. to 12:04 a.m." The project asks for the midpoint of each time range, so that code should become 00:02:00, and any code meaning "N/A" (such as people who don't commute) should become NA.

The project instructions allowed using an LLM for this part, and I used one to help write the helper_time() function below.The function looks up each code's label in the variable values I saved in the .rds file, splits the label into a start and end time, and returns the time halfway between them. I then use a for loop in get_census_data() that sends JWAP and JWDP through this helper and converts the other numeric variables with as.numeric().

Note : The helper_time() function needs to work with times, so the LLM suggested using the following two packages: lubridate, which reads text like "12:00 am" as a time, and hms, which stores the final result as just a time of day (like 00:02:00) without a date attached.

```{r}
library(lubridate) 
library(hms)

# helper: turn time codes into the midpoint time of each range
helper_time <- function(codes, item_lookup) {
  labels <- unlist(item_lookup)
  time_labels <- unname(labels[match(as.numeric(codes), as.numeric(names(labels)))])
  time_labels[grepl("N/A", time_labels)] <- NA
  time_labels <- gsub("\\.", "", time_labels)
  start <- parse_date_time(sub(" to .*", "", time_labels), "I:M p", quiet = TRUE)
  end <- parse_date_time(sub(".* to ", "", time_labels), "I:M p", quiet = TRUE)
  midpoint <- start + (end - start) / 2
  secs <- hour(midpoint) * 3600 + minute(midpoint) * 60 + second(midpoint)
  return(hms(seconds = secs))
}

get_census_data <- function(year = 2024,
                             num_vars = c("AGEP", "PWGTP"),
                             cat_vars = "SEX",
                             geo_level = "State",
                             geo_subset = NULL) { 
  
### check that a single valid year was given##############################
  if (length(year) != 1) {
    stop("you can only give one year at a time")
  }
  if (!(year %in% 2021:2024)) {
    stop("year must be 2021, 2022, 2023, or 2024")
  }
  
###check numeric variables################################################
   allowed_num_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  
   if (!all(num_vars %in% allowed_num_vars)) {
    stop(paste("numeric variables must be from:",
               paste(allowed_num_vars, collapse = ", ")))
  }
  if (!("PWGTP" %in% num_vars)) {
    stop("PWGTP must always be included")
  }
  if (length(setdiff(num_vars, "PWGTP")) < 1) {
    stop("you must include at least one numeric variable besides PWGTP")
  }
  
###check categorical variables###########################################
  allowed_cat_vars <- c("FER", "HHL", "SCH", "SCHL", "SEX") 
  
  if (length(cat_vars) < 1) {
    stop("you must include at least one categorical variable")
  }
  if (!all(cat_vars %in% allowed_cat_vars)) {
    stop(paste("categorical variables must be from:",
               paste(allowed_cat_vars, collapse = ", ")))
  }

 ### check geography level ###
  allowed_geo_levels <- c("Region", "Division", "State")

  if (length(geo_level) != 1) {
    stop("you can only choose one geography level")
  }
  if (!(geo_level %in% allowed_geo_levels)) {
    stop(paste("geo_level must be one of:",
               paste(allowed_geo_levels, collapse = ", ")))
  }

  ### set default subset if none given (NC and the areas it belongs to) ###
  if (is.null(geo_subset)) {
    if (geo_level == "State") {
      geo_subset <- "37"
    } else if (geo_level == "Division") {
      geo_subset <- "5"
    } else {
      geo_subset <- "3"
    }
  }

  ### check subset values against the saved variable values ###
  var_values <- readRDS("data/comb_list.rds")[[as.character(year)]]

  if (geo_level == "State") {
    geo_var <- ifelse("STATE" %in% names(var_values), "STATE", "ST")
  } else {
    geo_var <- toupper(geo_level)
  }
  allowed_codes <- names(var_values[[geo_var]]$item)

  if (!(length(geo_subset) == 1 && tolower(geo_subset) == "all")) {
    if (!all(geo_subset %in% allowed_codes)) {
      stop(paste("invalid", geo_level, "code. Allowed codes are:",
                 paste(allowed_codes, collapse = ", ")))
    }
  }

  ### turn the subset into the URL format ###
  if (length(geo_subset) == 1 && tolower(geo_subset) == "all") {
    geo_string <- "*"
  } else {
    geo_string <- paste(geo_subset, collapse = ",")
  }

  ### build the URL and query the API ###
  vars_string <- paste(c(num_vars, cat_vars), collapse = ",")
  url <- paste0("https://api.census.gov/data/",
                year,
                "/acs/acs1/pums?get=",
                vars_string,
                "&for=", tolower(geo_level), ":", geo_string,
                "&key=",
                api_key)
   
  data_rsp <- GET(url)
  data_all_tibble <- helper_tibble(data_rsp)
  
  ### convert numeric and time variables ###
  time_vars <- c("JWAP", "JWDP")

  for (v in num_vars) {
    if (v %in% time_vars) {
      data_all_tibble[[v]] <- helper_time(data_all_tibble[[v]], var_values[[v]]$item)
    } else {
      data_all_tibble[[v]] <- as.numeric(data_all_tibble[[v]])
    }
  }
  
  return(data_all_tibble)
 }
```

#### Fifth Test: Checking to see if the conversions worked. 

We request AGEP, JWAP, and PWGTP, then use str() to check that AGEP and PWGTP are now numeric and JWAP is stored as a time. Many JWAP values are NA because people who don't commute (such as children) have no arrival time, so we also look at the first few non-missing values to confirm they are real times of day. After running the test we can see that AGEP and PWGTP are now numeric, and the JWAP values are times such as 07:32:00, so the conversion worked as expected. The NA values belong to people with no commute time, like the 79 year old in the fourth row.

```{r}
test_time <- get_census_data(num_vars = c("AGEP", "JWAP", "PWGTP"))
head(test_time)
str(test_time)
```

Next we need to turn the categorical variables into labeled factors. Right now they come back from the API as codes, so SEX shows 1 or 2 and state shows a number like 37, which isn't very readable. The variable values I saved in the .rds file already list what each code means. I took inspiration from the helper_time function and came up with a simpler helper_factor() function that uses the saved variable value to replace each code with its label (for example, SEX 2 becomes "Female" and state 37 becomes "North Carolina/NC"). One problem is that codes aren't always written the same way in the data and in the saved variable value, for example "1" versus "01". 

To solve this issue the helper_factor function removes leading zeros from both the data and the saved variable value before matching them. I also apply this to the geography column so the state, region, or division names are readable. 

lastly I  add the "census" class to the tibble, which will let me write custom summary() and plot() functions later in the project. 

```{r}
# helper: turn categorical codes into a factor with labels
helper_factor <- function(codes, item_lookup) {
  labels <- unlist(item_lookup)

  # remove leading zeros from number codes so "01" and "1" match
  clean_codes <- function(x) {
    ifelse(grepl("^[0-9]+$", x), as.character(as.numeric(x)), x)
  }

  return(factor(clean_codes(codes),
                levels = clean_codes(names(labels)),
                labels = unname(labels)))
}

# helper: turn time codes into the midpoint time of each range
helper_time <- function(codes, item_lookup) {
  labels <- unlist(item_lookup)
  time_labels <- unname(labels[match(as.numeric(codes), as.numeric(names(labels)))])
  time_labels[grepl("N/A", time_labels)] <- NA
  time_labels <- gsub("\\.", "", time_labels)
  start <- parse_date_time(sub(" to .*", "", time_labels), "I:M p", quiet = TRUE)
  end <- parse_date_time(sub(".* to ", "", time_labels), "I:M p", quiet = TRUE)
  midpoint <- start + (end - start) / 2
  secs <- hour(midpoint) * 3600 + minute(midpoint) * 60 + second(midpoint)
  return(hms(seconds = secs))
}

get_census_data <- function(year = 2024,
                             num_vars = c("AGEP", "PWGTP"),
                             cat_vars = "SEX",
                             geo_level = "State",
                             geo_subset = NULL) { 
  
### check that a single valid year was given##############################
  if (length(year) != 1) {
    stop("you can only give one year at a time")
  }
  if (!(year %in% 2021:2024)) {
    stop("year must be 2021, 2022, 2023, or 2024")
  }
  
###check numeric variables################################################
   allowed_num_vars <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  
   if (!all(num_vars %in% allowed_num_vars)) {
    stop(paste("numeric variables must be from:",
               paste(allowed_num_vars, collapse = ", ")))
  }
  if (!("PWGTP" %in% num_vars)) {
    stop("PWGTP must always be included")
  }
  if (length(setdiff(num_vars, "PWGTP")) < 1) {
    stop("you must include at least one numeric variable besides PWGTP")
  }
  
###check categorical variables###########################################
  allowed_cat_vars <- c("FER", "HHL", "SCH", "SCHL", "SEX") 
  
  if (length(cat_vars) < 1) {
    stop("you must include at least one categorical variable")
  }
  if (!all(cat_vars %in% allowed_cat_vars)) {
    stop(paste("categorical variables must be from:",
               paste(allowed_cat_vars, collapse = ", ")))
  }

 ### check geography level ###
  allowed_geo_levels <- c("Region", "Division", "State")

  if (length(geo_level) != 1) {
    stop("you can only choose one geography level")
  }
  if (!(geo_level %in% allowed_geo_levels)) {
    stop(paste("geo_level must be one of:",
               paste(allowed_geo_levels, collapse = ", ")))
  }

  ### set default subset if none given (NC and the areas it belongs to) ###
  if (is.null(geo_subset)) {
    if (geo_level == "State") {
      geo_subset <- "37"
    } else if (geo_level == "Division") {
      geo_subset <- "5"
    } else {
      geo_subset <- "3"
    }
  }

  ### check subset values against the saved variable values ###
  var_values <- readRDS("data/comb_list.rds")[[as.character(year)]]

  if (geo_level == "State") {
    geo_var <- ifelse("STATE" %in% names(var_values), "STATE", "ST")
  } else {
    geo_var <- toupper(geo_level)
  }
  allowed_codes <- names(var_values[[geo_var]]$item)

  if (!(length(geo_subset) == 1 && tolower(geo_subset) == "all")) {
    if (!all(geo_subset %in% allowed_codes)) {
      stop(paste("invalid", geo_level, "code. Allowed codes are:",
                 paste(allowed_codes, collapse = ", ")))
    }
  }

  ### turn the subset into the URL format ###
  if (length(geo_subset) == 1 && tolower(geo_subset) == "all") {
    geo_string <- "*"
  } else {
    geo_string <- paste(geo_subset, collapse = ",")
  }

  ### build the URL and query the API ###
  vars_string <- paste(c(num_vars, cat_vars), collapse = ",")
  url <- paste0("https://api.census.gov/data/",
                year,
                "/acs/acs1/pums?get=",
                vars_string,
                "&for=", tolower(geo_level), ":", geo_string,
                "&key=",
                api_key)
   
  data_rsp <- GET(url)
  data_all_tibble <- helper_tibble(data_rsp)
  
  ### convert numeric and time variables ###
  time_vars <- c("JWAP", "JWDP")

  for (v in num_vars) {
    if (v %in% time_vars) {
      data_all_tibble[[v]] <- helper_time(data_all_tibble[[v]], var_values[[v]]$item)
    } else {
      data_all_tibble[[v]] <- as.numeric(data_all_tibble[[v]])
    }
  }
  
    ### convert categorical variables to labeled factors ###
  for (v in cat_vars) {
    data_all_tibble[[v]] <- helper_factor(data_all_tibble[[v]], var_values[[v]]$item)
  }

  # also label the geography column (state, region, or division)
  geo_col <- tolower(geo_level)
  data_all_tibble[[geo_col]] <- helper_factor(data_all_tibble[[geo_col]],
                                              var_values[[geo_var]]$item)

  ### add the census class (layer E) ###
  class(data_all_tibble) <- c("census", class(data_all_tibble))
  
  return(data_all_tibble)
 }
```
####Sixth Test: Checking to see if the categorical variables were converted correctly.

We request SEX and SCHL, then check that head() shows labels instead of codes, that str() lists these columns as factors, and that class() includes "census." SCHL is a good test because its codes may be written with leading zeros, so it shows whether the matching between the data and the dictionary works.

```{r}
test_factor <- get_census_data(cat_vars = c("SEX", "SCHL"))
head(test_factor)
str(test_factor)
class(test_factor)
```

SEX, SCHL, and state are now factors with readable labels, such as "Female," "Regular high school diploma," and "North Carolina/NC," so the codes matched the dictionary correctly!! and importantly for part two The class() output shows "census" first, followed by the usual tibble classes. 

###Combining Years



#### Test

##Part 2

###Writing Generic Functions for Summarizing

###Numeric Summaries

####Test 

###Graphical Summaries 

####Test

##Investigation
##Conclusion 

