---
title: "Project 1 - Maab Aldulimy"
format: html
editor: visual
---

# Project Goal

First Project in ST558 the goal is to write functions that will manipulate and process data sets that come from a census API. The generic functions should automatically summarize and plot certain returned data. Lastly, a document via quarto will be written to describe the thought process and give examples of using the functions.

## Data

The data for this project will be acquired from the Public Use Microdata Sample (PUMS) Census API. This API gives pseudo person level data. An API was requested and acquired from the website 


```{r}
# loading in needed packages 

library(dplyr)
library(httr)
library(jsonlite)
library(ggplot2)

# masking API
api_key <- Sys.getenv("CENSUS_API_KEY")

if (api_key == "") {
  stop("CENSUS_API_KEY has not been set.")
}
```

For this section, we connect to the census API

We use GET() from the httr package to return information about the variables in the census. our aim at this point is to make sure the url works and the we can access the information we want in this case the items under the variable sex. 
```{r}
url <- paste0("https://api.census.gov/data/2024/acs/acs1/pums/variables.json")

variables_f <- GET(url)
vars_2024<-fromJSON(content(variables_f,"text"))

#testing to see if i can access male/female 
vars_2024$variables$SEX$values$item


```

Make a character vector of the 15 variable names ( AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP, PWGTP, FER, HHL, SCH, SCHL, SEX, ST (for 2021, 2022)/STATE (for 2023, 2024), REGION, and DIVISION) and subset vars_2024$variables with single brackets.

```{r}
 vars_2024_sub<-vars_2024$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST","STATE","REGION","DIVISION")]

names(vars_2024_sub)
length(vars_2024_sub)
```
```{r}
wanted_vars<-c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST","STATE","REGION","DIVISION")

names(vars_2024$variables)
wanted_vars %in% names(vars_2024$variables)
vars_to_keep <- wanted_vars[wanted_vars %in% names(vars_2024$variables)]
vars_to_keep 

vars_2024_sub<-vars_2024$variables[vars_to_keep]
names(vars_2024_sub)
length(vars_2024_sub)
```

